---
title: "Differences between disparity through time in the `geiger` and `dispRity` packages."
author: "Thomas Guillerme"
date: "`r Sys.Date()`"
output:
  html_document:
    fig_width: 12
    fig_height: 6
---


```{r}
library(dispRity)
library(geiger)

## Setting up the data
geo  <- get(data(geospiza))
tree <- geo$phy
data <- geo$dat
cleaned <- clean.data(data, tree)
tree <- cleaned$tree
data <- cleaned$data
tree <- makeNodeLabel(tree, prefix = "n")
tree$root.time <- max(tree.age(tree)$age)

## running ancestral state estimations
ace_data <- multi.ace(data, tree)
full_data <- rbind(data, ace_data)

## the metric from geiger
average.sq <- function(X) mean(pairwise.dist(X)^2)
```

## `geiger::dtt` vs. `dispRity::dtt.dispRity`

First let's look at the two implementations of `dtt`: `geiger::dtt` and `disprity::dtt.dispRity`.
The second one is just a wrapper for the first one with a modified modularity for the metrics arguments.
Basically `geiger::dtt` is limited three disparity metrics (provided by the `index` argument) while the `dispRity` version generalises that to any metric valid for the `dispRity` package.
For example, it could work with the `c(sum, variances)` one or, here, the `average.sq` one which is defined above the same way as `"avg.sq"`.

```{r}
## dtt from geiger
dtt_geiger <- geiger::dtt(phy = tree, data = data, nsim = 20, plot = TRUE, index = "avg.sq")

## dtt from dispRity (a wrapper for geiger:dtt but with modularity for metrics in the dispRity format)
dtt_dispRity <- dispRity::dtt.dispRity(data = data, tree = tree, metric = average.sq)

## Adding the dispRity wrapper version of dtt
lines(x = dtt_geiger$times, y = dtt_dispRity$dtt, col = "green", lwd = 1, lty = 1)
legend("topleft", lwd = 2, col = c("black", "green"),
       legend = c("DTT (geiger::dtt)", "disparity through time\n(dispRity::chrono.subsets)"))
```

They're exactly the same, as expected.

## disparity-through-time (`dtt`) vs. disparity through time (`chrono.subsets`)

Now let's see the difference between the DTT version and the one in implemented in `chrono.subsets` in `dispRity` (with a `"proximity"` model just for illustration, note that the results are the same for each different model).

```{r}
## dtt from geiger
dtt_geiger <- geiger::dtt(phy = tree, data = data, nsim = 20, plot = TRUE, index = "avg.sq")

## disparity through time from dispRity
## Get the time slices to correspond to the ones in DTT
time_slices <- ((1-dtt_dispRity$times) * tree$root.time)[-1]
time_subsets <- chrono.subsets(full_data, tree = tree, method = "continuous", model = "proximity", time = time_slices)
## Get the disparity through time
disparity_through_time <- dispRity(time_subsets, metric = average.sq)
## Scale the disparity through time
disparity_through_time <- unlist(get.disparity(disparity_through_time))
disparity_through_time <- disparity_through_time[-1]/max(disparity_through_time[-1])

## And here is the difference (remove the two first time points that don't have enough data for chrono.subsets)
lines(x = dtt_geiger$times[-c(1,2)], y = disparity_through_time, col = "orange", lwd = 2)
legend("topleft", lwd = 2, col = c("black", "orange"),
       legend = c("DTT (geiger::dtt)", "disparity through time\n(dispRity::chrono.subsets)"))
```

The results are pretty much very different here.
How comes?
This is because of the difference in how either methods are measuring disparity through time.
 
 * `dispRity::chrono.subsets` are actually measuring disparity through time in a commonly understood way: the x axis represents time and the y axis represents disparity at that point in time. I say "commonly understood" because this matches more to the way we are measuring a variable through time (e.g. population size through time for a region, height through time for an individual, available budget through time, etc...)
 * `geiger::dtt` measures disparity through time in the sense of clade disparity through time. In other words, for each node, it's looking at the proportion of disparity of the clade (i.e. the node's descendants) to the overall disparity. It then calculates at each point (the age of the nodes), the average disparity in the subclades present at that time.

Here's a way to visualise the decomposition of what `geiger::dtt` is doing.

```{r}
## Creating the groups of all nodes and their descendants.
clade_groups <- custom.subsets(data, group = tree)
## Note here that providing group = tree generates a group for each node
## i.e. node 1 (n1) contains the 13 species (the root of the tree),
## node 2 (n2) 12 species, n3, 11 sp, etc.
size.subsets(clade_groups)

## We can then calculate the disparity for these groups
clade_disparity <- dispRity(clade_groups, metric = average.sq)
disparity_per_clade <- unlist(get.disparity(clade_disparity))
## And scale these values by the biggest disparity value
disparity_per_clade <- disparity_per_clade/max(disparity_per_clade) 

## And here is what it looks like
dtt_geiger <- geiger::dtt(phy = tree, data = data, nsim = 20, plot = TRUE, index = "avg.sq")
lines(x = dtt_geiger$times[-c(1)], y = disparity_per_clade, col = "blue", lwd = 2)

## Adding the legend
legend("topleft", lwd = 2, col = c("black", "blue"),
       legend = c("DTT (geiger::dtt)", "disparity per clade (dispRity)"))
```

Note that here the results are a bit similar that DTT but still quiet different.

This is because DTT re-adjust these values to scale them and calculate the average at each time points (which are the node ages).
Looking at a reproduction of the `geiger::dtt` code, this is what it looks like

```{r}
## Get the node ages for as the time slices
node_ages <- tree.age(tree)$ages[-c(1:Ntip(tree))]
names(node_ages) <- 1:Nnode(tree) + Ntip(tree)

## Finding the ancestral stem age for each node for averaging laterone
ancestor_ages <- numeric()
ancestor_ages <- unname(node_ages[1])
for(node in 2:tree$Nnode) {
       ## Find this node's ancestor
       node_anc_edge <- which(tree$edge[, 2] == node+Ntip(tree))
       ## Getting the node ancestor's age
       ancestor_ages[node] <- node_ages[as.numeric(names(node_ages)) == tree$edge[node_anc_edge, 1]]
}

## Sort the node depths sequentially
sorted_ages <- sort(node_ages, decreasing = TRUE)

## Calculate disparity per clade
disparity <- unlist(get.disparity(dispRity(custom.subsets(data, tree), metric = average.sq)))

## Disparity at the root
dtt_results <- numeric()
dtt_results[1] <- disparity[1]

## Disparity for the other nodes (average per time slice)
for(i in 2:length(sorted_ages)) {
       ## Check which lineages are present at the time of interest 
       x <- disparity[ancestor_ages >= sorted_ages[i-1] & node_ages < sorted_ages[i-1]]
       ## ancetral node ages is bigger or equal than previous ltt AND node depth is smaller than previous ltt
       if(length(x) == 0) {
              ## If none, the disparity is 0
              dtt_results[i] <- 0
       } else {
              ## Else, it's the average disparity of the clades present at that time
              dtt_results[i] <- mean(x)
       }
}

## The last disparity value is set to 0
dtt_results[length(sorted_ages)+1] <- 0

## And the results are scaled relative to the first observation (the disparity of the whole clade)
dtt_results <- dtt_results/dtt_results[1]

## Adding the different disparity scores
dtt_geiger <- geiger::dtt(phy = tree, data = data, nsim = 20, plot = TRUE, index = "avg.sq")
lines(x = dtt_geiger$times, y = dtt_results, col = "orange", lwd = 1)

## You can compare that to the more naive approach highlighted above
## where the disparity per clade is just plotted through time:
raw_disparity <- disparity/disparity[1]
raw_disparity <- c(raw_disparity, 0)

lines(x = dtt_geiger$times, y = raw_disparity, col = "blue", lwd = 2)

## Adding the legend
legend("topleft", lwd = 2, col = c("black", "orange", "blue"),
       legend = c("DTT (geiger::dtt)", "average disparity per clade", "raw disparity per clade"))
```

<!-- Refine the algorithm with slice.tree(tree, age = sort(node_ages)[3], model = "ACCTRAN")
? -->