---
title: "Getting started with dispRity"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
bibliography: [../References.bib, ../packages.bib]
output:
  html_document: default
  pdf_document: default
---

# Getting started with `dispRity`

## What sort of data does `dispRity` work with?

**Any matrix object in `R`.**
Disparity can be estimated from pretty much any matrix as long as rows represent the elements and columns the dimensions.
These matrices can be observations, pairwise differences between elements, ordinations, etc...

> Note that whether the matrix is an ordinated one or of any other type, it is always considered as the multidimensional space to analysis and no corrections are applied to it (e.g. if ordinated and with negative eigen values, no correction is used). Always make sure the input is the right multidimensional space to analyse!

## Ordinated matrices

Classically, when a high number of variables is used, disparity is calculated from ordinated matrices.
These can be any type of ordinations (PCO, PCA, PCoA, MDS, etc.) as long as elements are the rows (taxa, countries, field experiments) and the dimensions are the columns.

### Ordination matrices from `geomorph`

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if geomorph is installed BUILD ONLY
if(!require(geomorph)) install.packages("geomorph")
```

You can also easily use data from `geomorph` using the `geomorph.ordination` function.
This function simply takes Procrustes aligned data and performs an ordination:

```{r}
require(geomorph)

## Loading the plethodon dataset
data(plethodon)

## Performing a Procrustes transform on the landmarks
procrustes <- gpagen(plethodon$land, PrinAxes = FALSE, print.progress = FALSE)

## Ordinating this data
geomorph.ordination(procrustes)[1:5,1:5]
```

Options for the ordination (from `?prcomp`) can be directly passed to this function to perform customised ordinations.
Additionally you can give the function a `geomorph.data.frame` object.
If the latter contains sorting information (i.e. factors), they can be directly used to make a customised `dispRity` object [customised `dispRity` object](#customised-subsets)!

```{r}
## Using a geomorph.data.frame
geomorph_df <- geomorph.data.frame(procrustes,
     species = plethodon$species, site = plethodon$site)

## Ordinating this data and making a dispRity object
geomorph.ordination(geomorph_df)
```

More about these `dispRity` objects below!

### Ordination matrices from `Claddis` {#Claddis-ordination}

> This functionality is not available on the CRAN version of the package (`r version_CRAN`) due to CRAN compatibility issues see [to CRAN or not to CRAN?](#noCRAN) for more information and on how to install the GitHub (more complete) version of the package (`r version_release`).

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## Checking if Claddis is installed BUILD ONLY
if(!require(Claddis)) install.packages("Claddis")
```

`dispRity` package can easily take data from `Claddis` using the `Claddis.ordination` function.
For this, simply input a matrix in the `Claddis` format to the function and it will automatically calculate and ordinate the distances among taxa:

```{r}
require(Claddis)

## Ordinating the example data from Claddis
Claddis.ordination(Michaux1989) 
```

Note that several options are available, namely which type of distance should be computed.
See more info in the function manual (`?Claddis.ordination`).
Alternatively, it is of course also possible to manual calculate the ordination matrix using the functions `Claddis::MorphDistMatrix` and `stats::cmdscale`.

### Other kinds of ordination matrices

If you are not using the packages mentioned above (`Claddis` and `geomorph`) you can easily make your own ordination matrices by using the following functions from the `stats` package.
Here is how to do it for the following types of matrices:

 * Multivariate matrices (principal components analysis; PCA)

```{r}
## A multivariate matrix
head(USArrests)

## Ordinating the matrix using `prcomp` 
ordination <- prcomp(USArrests)

## Selecting the ordinated matrix
ordinated_matrix <- ordination$x
head(ordinated_matrix)
```
  This results in a ordinated matrix with US states as elements and four dimensions (PC 1 to 4). For an alternative method, see the `?princomp` function.

 * Distance matrices (classical multidimensional scaling; MDS)

```{r}
## A matrix of distances between cities
str(eurodist)

## Ordinating the matrix using cmdscale() with k = 5 dimensions 
ordinated_matrix <- cmdscale(eurodist, k = 5)
head(ordinated_matrix)
```

  This results in a ordinated matrix with European cities as elements and five dimensions.

Of course any other method for creating the ordination matrix is totally valid, you can also not use any ordination at all!
The only requirements for the `dispRity` functions is that the input is a matrix with elements as rows and dimensions as columns.

## Performing a simple dispRity analysis {#simpleanalysis}

Two `dispRity` functions allow users to run an analysis pipeline simply by inputting an ordination matrix.
These functions allow users to either calculate the disparity through time (`dispRity.through.time`) or the disparity of user-defined groups (`dispRity.per.group`).

**IMPORTANT**

Note that `disparity.through.time` and `disparity.per.group` are wrapper functions (i.e. they incorporate lots of other functions) that allow users to run a basic disparity-through-time, or disparity among groups, analysis without too much effort.
As such they use a lot of default options. 
These are described in the help files for the functions that are used to make the wrapper functions, and not described in the help files for `disparity.through.time` and `disparity.per.group`. 
These defaults are good enough for **data exploration**, but for a proper analysis you should consider the **best parameters for your question and data**. 
For example, which metric should you use?
How many bootstraps do you require?
What model of evolution is most appropriate if you are time slicing?
Should you rarefy the data?
See [`chrono.subsets`](#time-slicing), [`custom.subsets`](#customised-subsets), [`boot.matrix`](#bootstraps-and-rarefactions) and [`dispRity.metric`](#disparity-metrics) for more details of the defaults used in each of these functions.
Note that any of these default arguments can be changed within the `disparity.through.time` or `disparity.per.group` functions.

### Example data

To illustrate these functions, we will use data from @beckancient2014.
This dataset contains an ordinated matrix of 50 discrete characters from mammals (`BeckLee_mat50`), another matrix of the same 50 mammals and the estimated discrete data characters of their descendants (thus 50 + 49 rows, `BeckLee_mat99`), a dataframe containing the ages of each taxon in the dataset (`BeckLee_ages`) and finally a phylogenetic tree with the relationships among the 50 mammals (`BeckLee_tree`).

```{r, fig.width=7, fig.height=7}
## Loading the ordinated matrices
data(BeckLee_mat50)
data(BeckLee_mat99)

## The first five taxa and dimensions of the 50 taxa matrix
head(BeckLee_mat50[, 1:5])

## The first five taxa and dimensions of the 99 taxa + ancestors matrix
BeckLee_mat99[c(1, 2, 98, 99), 1:5]

## Loading a list of first and last occurrence dates for the fossils
data(BeckLee_ages)
head(BeckLee_ages)

## Loading and plotting the phylogeny
data(BeckLee_tree)
plot(BeckLee_tree, cex = 0.8) 
axisPhylo(root = 140)
nodelabels(cex = 0.5)
```

Of course you can use your own data as detailed in the [previous chapter](#What-sort-of-data-does-dispRity-work-with).

### Disparity through time

The `dispRity.through.time` function calculates disparity through time, a common analysis in palaeontology.
This function (and the following one) uses an analysis pipeline with a lot of default parameters to make the analysis as simple as possible. 
Of course all the defaults can be changed if required, more on this later.

For a disparity through time analysis, you will need:
  
  * An ordinated matrix (we covered that above)
  * A phylogenetic tree: this must be a `phylo` object (from the `ape` package) and needs a `root.time` element. To give your tree a root time (i.e. an age for the root), you can simply do\\ `my_tree$root.time <- my_age`.
  * The required number of time subsets (here `time = 3`)
  * Your favourite disparity metric (here the sum of variances)

Using the Beck and Lee (2014) data described [above](#example-data):

```{r}
## Measuring disparity through time
disparity_data <- dispRity.through.time(BeckLee_mat50, BeckLee_tree,
                                        time = 3, metric = c(sum, variances))
```

This generates a `dispRity` object (see [here](#guts) for technical details).
When displayed, these `dispRity` objects provide us with information on the operations done to the matrix:

```{r}
## Print the disparity_data object
disparity_data
```

We asked for three subsets (evenly spread across the age of the tree), the data was bootstrapped 100 times (default) and the metric used was the sum of variances.

We can now summarise or plot the `disparity_data` object, or perform statistical tests on it (e.g. a simple `lm`): 


```{r, fig.width=7, fig.height=7}
## Summarising disparity through time
summary(disparity_data)

## Plotting the results
plot(disparity_data, type = "continuous")

## Testing for an difference among the time bins
disp_lm <- test.dispRity(disparity_data, test = lm, comparisons = "all")
summary(disp_lm)
```

Please refer to the [specific tutorials](#specific-tutorial) for (much!) more information on the nuts and bolts of the package.
You can also directly explore the specific function help files within R and navigate to related functions.

### Disparity among groups

The `dispRity.per.group` function is used if you are interested in looking at disparity among groups rather than through time.
For example, you could ask if there is a difference in disparity between two groups?

To perform such an analysis, you will need:
 
 * An matrix with rows as elements and columns as dimensions (always!)
 * A list of group members: this list should be a list of numeric vectors or names corresponding to the row names in the matrix. For example `list("a" = c(1,2), "b" = c(3,4))` will create a group _a_ containing elements 1 and 2 from the matrix and a group _b_ containing elements 3 and 4. Note that elements can be present in multiple groups at once.
 * Your favourite disparity metric (here the sum of variances)

Using the @beckancient2014 data described [above](#example-data):

```{r}
## Creating the two groups (crown versus stem) as a list
mammal_groups <- crown.stem(BeckLee_tree, inc.nodes = FALSE)

## Measuring disparity for each group
disparity_data <- dispRity.per.group(BeckLee_mat50, group = mammal_groups,
                                     metric = c(sum, variances))
```

We can display the disparity of both groups by simply looking at the output variable (`disparity_data`) and then summarising the `disparity_data` object and plotting it, and/or by performing a statistical test to compare disparity across the groups (here a Wilcoxon test).

```{r, fig.width=7, fig.height=7}
## Print the disparity_data object
disparity_data

## Summarising disparity in the different groups
summary(disparity_data)

## Plotting the results
plot(disparity_data)

## Testing for a difference between the groups
test.dispRity(disparity_data, test = wilcox.test, details = TRUE)
```