---
title: "Data simulation features"
author: "Thomas Guillerme (guillert@tcd.ie) and Natalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

# Making stuff up!

The `dispRity` package also offers some advanced data simulation features to allow to test hypothesis, explore ordinate-spaces or metrics properties or simply playing around with data!
All the following functions are based on the same modular architecture of the package and therefore can be used with most of the functions of the package.

## Simulating discrete morphological data


The function `sim.morpho` allows to simulate discrete morphological data matrices (sometimes referred to as "cladistic" matrices).
It allows to evolve multiple discrete characters on a given phylogenetic trees, given different models, rates, and states.
It even allows to include proper inapplicable data to make datasets as messy as in real life!

In brief, the function `sim.morpho` takes a phylogenetic tree, the number of required characters, the evolutionary model, and a function from which to draw the rates.
The package also contains a function for quickly checking the matrix's phylogenetic signal (as defined in systematics not phylogenetic comparative methods) using parsimony.
The methods are described in details below


```{r}
set.seed(3)
## Simulating a starting tree with 15 taxa as a random coalescent tree
my_tree <- rcoal(15)

## Generating a matrix with 100 characters (85% binary and 15% three state) and
## an equal rates model with a gamma rate distribution (0.5, 1) with no 
## invariant characters.
my_matrix <- sim.morpho(tree = my_tree, characters = 100, states = c(0.85,
    0.15), rates = c(rgamma, 0.5, 1), invariant = FALSE)

## The first few lines of the matrix
my_matrix[1:5, 1:10]

## Checking the matrix properties with a quick Maximum Parsimony tree search
check.morpho(my_matrix, my_tree)
```

Note that this example produces a tree with a great consistency index and an identical topology to the random coalescent tree!
Nearly too good to be true...

### A more detailed description

The protocol implemented here to generate discrete morphological matrices is based on the ones developed in [@GuillermeCooper;@OReilly2016;@puttick2017uncertain;OReilly2017].

 * The first `tree` argument will be the tree on which to "evolve" the characters and therefore requires branch length.
 You can generate quick and easy random Yule trees using `ape::rtree(number_of_taxa)` but I would advise to use more realistic trees for more realistic simulations based on more realistic models (really realistic then) using the function `tree.bd` from the [`diversitree`](http://www.zoology.ubc.ca/prog/diversitree/) package [@fitzjohndiversitree2012].
 * The second argument, `character` is the number of characters. Pretty straight forward.
 * The third, `states` is the proportion of characters states above two (yes, the minimum number of states is two). This argument intakes the proportion of *n*-states characters, for example `states = c(0.5,0.3,0.2)` will generate 50% of binary-state characters, 30% of three-state characters and 20% of four-state characters. There is no limit in the number of state characters proportion as long as the total makes up 100%. 
 * The forth, `model` is the evolutionary model for generating the character(s). More about this below.
 * The fifth and sixth, `rates` and `substitution` are the model parameters described below as well.
 * Finally, the two logical arguments, are self explanatory: `invariant` whether to allow invariant characters (i.e. characters that don't change) and `verbose` whether to print the simulation progress on your console.

#### Available evolutionary models

There are currently three evolutionary models implemented in `sim.morpho` but more will come in the future.
Note also that they allow fine tuning parameters making them pretty plastic!

 * `"ER"`: this model allows any number of character states and is based on the Mk model [@lewisa2001]. It assumes a unique overall evolutionary rate equal substitution rate between character states. This model is based on the `ape::rTraitDisc` function.
 * `"HKY"`: this is binary state character model based on the molecular HKY model [@HKY85]. It uses the four molecular states (A,C,G,T) with a unique overall evolutionary rate and a biased substitution rate towards transitions (A <-> G or C <-> T) against transvertions (A <-> C and G <-> T). After evolving the nucleotide, this model transforms them into binary states by converting the purines (A and G) into state 0 and the pyrimidines (C and T) into state 1. This method is based on the `phyclust::seq.gen.HKY` function and was first proposed by @OReilly2016.
 * `"MIXED"`: this model uses a random (uniform) mix between both the `"ER"` and the `"HKY"` models.

The models can take the following parameters:
(1) `rates` is the evolutionary rate (i.e. the rate of changes along a branch: the evolutionary speed) and (2) `substitution` is the frequency of changes between one state or another.
For example if a character can have high probability of changing (the *evolutionary* rate) with, each time a change occurs a probability of changing from state *X* to state *Y* (the *substitution* rate).
Note that in the `"ER"` model, the substitution rate is ignore because... by definition this (substitution) rate is equal!

The parameters arguments `rates` and `substitution` takes a distributions from which to draw the parameters values for each character.
For example, if you want an `"HKY"` model with an evolutionary rate (i.e. speed) drawn from a uniform distribution bounded between 0.001 and 0.005, you can define it as `rates = c(runif, min = 0.001, max = 0.005)`, `runif` being the function for random draws from a uniform distribution and `max` and `min` being the distribution parameters.
These distributions should always be passed in the format `c(random_distribution_function, distribution_parameters)` with the names of the distribution parameters arguments.

#### Checking the results

An additional function, `check.morpho` runs a quick Maximum Parsimony tree search using the `phangorn` parsimony algorithm.
It quickly calculates the parsimony score, the consistency and retention indices and, if a tree is provided (e.g. the tree used to generate the matrix) it calculates the Robinson-Foulds distance between the most parsimonious tree and the provided tree to determine how different they are.

#### Adding inapplicable characters

Once a matrix is generated, it is possible to apply inapplicable characters to it for increasing realism!
Inapplicable characters are commonly designated as `NA` or simply `-`.
They differ from missing characters `?` in their nature by being inapplicable rather than unknown.
For example, considering a binary character defined as "color of the tail" with the following states "blue" and "red"; on a taxon with no tail, the character should be coded as inapplicable ("`-`") since the state of the character "color of tail" is *known*: it's neither "blue" or "red", it's just not there!
It contrasts with coding it as missing ("`?`" - also called as ambiguous) where the state is *unknown*, for example, the taxon of interest is a fossil where the tail has no colour preserved or is not present at all due to bad conservation!

This type of characters can be added to the simulated matrices using the `apply.NA` function/
It takes, as arguments, the `matrix`, the source of inapplicability (`NAs` - more below), the `tree` used to generate the matrix and the two same `invariant` and `verbose` arguments as defined above.
The `NAs` argument allows two types of sources of inapplicability:

 * `"character"` where the inapplicability is due to the character (e.g. coding a character tail for species with no tail).
 In practice, the algorithm chooses a character *X* as the underlying character (e.g. "presence and absence of tail"), arbitrarily chooses one of the states as "absent" (e.g. 0 = absent) and changes in the next character *Y* any state next to character *X* state 0 into an inapplicable token ("`-`").
 This simulates the inapplicability induced by coding the characters (i.e. not always biological).
 * `"clade"` where the inapplicability is due to evolutionary history (e.g. a clade loosing its tail).
 In practice, the algorithm chooses a random clade in the tree and a random character *Z* and replaces the state of the taxa present in the clade by the inapplicable token ("`-`").
 This simulates the inapplicability induced by evolutionary biology (e.g. the lose of a feature in a clade).

To apply these sources of inapplicability, simply repeat the number of inapplicable sources for the desired number of characters with inapplicable data.

```{r}
## Generating 5 "character" NAs and 10 "clade" NAs
my_matrix_NA <- apply.NA(my_matrix, tree = my_tree,
                         NAs = c(rep("character", 5), rep("clade", 10)))

## The first few lines of the resulting matrix
my_matrix_NA[1:10, 1:10]
```
<!--TG: Add inapplicable paper here -->

### Parameters for a realistic(ish) matrix

There are many parameters that can create a "realistic" matrix (i.e. not too different from the input tree with a consistency and retention index close to what is seen in the literature) but because of the randomness of the matrix generation not all end up creating "good" matrices.
The following parameters however, seem to generate fairly "realist" matrices with a starting coalescent tree, equal rates model with 0.85 binary characters and 0.15 three state characters, a gamma distribution with a shape parameter ($\alpha$) of 5 and no scaling ($\beta$ = 1) with a rate of 100.

```{r}
set.seed(0)
## tree
my_tree <- rcoal(15)
## matrix
morpho_mat <- sim.morpho(my_tree, characters = 100, model = "ER",
    rates = c(rgamma, rate = 100, shape = 5), invariant = FALSE)
check.morpho(morpho_mat, my_tree)
```

## Simulating multidimensional spaces
Another way to simulate data is to directly simulate an ordinated space with the `space.maker` function.
This function allows users to simulate multidimensional spaces with a certain number of properties.
It takes as arguments the number of elements (data points) and dimensions to create the space but also permits more fine tuning in the data simulation: it is possible to give a specific distribution to each dimension, provide a correlation matrix to link the dimensions or even provide an *a priori* distribution of the variance for each distributions!
<!-- ooh fancy! -->

[@diaz2016global]


```{r, fig.width=8, fig.height=8, eval=TRUE}
## Graphical options
op <- par(mfrow = (c(2, 2)), bty = "n")
## Visualising 3D spaces
require(scatterplot3d)

## A cube space
scatterplot3d(space.maker(2500, 3, runif), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Uniform cube space")

## A plane space
scatterplot3d(space.maker(2500, 3, c(runif, runif, runif),
              arguments = list(list(min = 0, max = 0), NULL, NULL)), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Uniform plane space")

## An ellipsoid space (=a spheric space with correlation)
cor_matrix <- matrix(cbind(1, 0.8, 0.2, 0.8, 1, 0.7, 0.2, 0.7, 1), nrow = 3)
scatterplot3d(space.maker(2500, 3, rnorm, cor.matrix = cor_matrix), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Normal ellipsoid space")

## A cylindrical space with a decreasing variance per axis
scatterplot3d(space.maker(2500, 3, c(rnorm, rnorm, runif),
              scree = c(0.7, 0.2, 0.1)), pch = 20,
              xlab = "", ylab = "", zlab = "",
              main = "Normal cylindrical space")

## Resetting the graphic parameters
par(op)
```
