--- 
title: "dispRity manual"
author: "Thomas Guillerme (guillert@tcd.ie),\nMark Puttick (marknputtick@gmail.com) and\nNatalie Cooper (natalie.cooper@nhm.ac.uk)"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
bibliography: [../References.bib, ../packages.bib]
link-citations: yes
github-repo: TGuillerme/dispRity
description: "dispRity R package vignette"
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, eval = TRUE, echo = FALSE}
## master version
version_raw <- RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/master/DESCRIPTION")
version_master <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
version_master <- 1.5

## release version
version_raw <- RCurl::getURL("https://raw.githubusercontent.com/TGuillerme/dispRity/release/DESCRIPTION")
version_release <- strsplit(strsplit(version_raw, split = "Version: ")[[1]][2], split = "\nDate:")[[1]][1]
version_release <- 1.5

## CRAN version
version_raw <- RCurl::getURL("https://cran.r-project.org/web/packages/dispRity/index.html")
version_CRAN <- strsplit(strsplit(version_raw, split = "\n<td>Version:</td>\n<td>")[[1]][2], split = "</td>\n</tr>\n<tr>\n<td>Depends:")[[1]][1]
version_CRAN <- 1.5
```

# `dispRity`

This is a package for measuring disparity (aka multidimensional space occupancy) in `R`.
It allows users to summarise matrices as representations as multidimensional spaces into a single value or distribution describing a specific aspect of this multidimensional space (the disparity).
Multidimensional spaces can be ordinated matrices from MDS, PCA, PCO, PCoA but the package is *not* restricted to any type of matrices!
This manual is based on the version `r version_release`.

## What is `dispRity`?
This is a modular package for measuring disparity in `R`.
It allows users to summarise ordinated matrices (e.g. MDS, PCA, PCO, PCoA) to perform some multidimensional analysis.
Typically, these analysis are used in palaeobiology and evolutionary biology to study the changes in morphology through time.
However, there are many more applications in ecology, evolution and beyond.

### Modular?
Because their exist a multitude of ways to measure disparity, each adapted to every specific question, this package uses an easy to modify modular architecture.
In coding, each module is simply a function or a modification of a function that can be passed to the main functions of the package to tweak it to your proper needs!
In practice, you will notice throughout this manual that some function can take other functions as arguments: the modular architecture of this package allows you to use any function for these arguments (with some restrictions explained for each specific cases).
This will allow you to finely tune your multidimensional analysis to the needs of your specific question!

## Installing and running the package

You can install this package easily, directly from the CRAN:

```{r, eval = FALSE}
install.packages("dispRity")
```

Alternatively, for the most up to data version and some functionalities not compatible with the CRAN, you can use the package through GitHub using `devtool` (see [to CRAN or not to CRAN?](#noCRAN) for more details):

```{r, eval=FALSE}
## Checking if devtools is already installed
if(!require(devtools)) install.packages("devtools")

## Installing the latest released version directly from GitHub
install_github("TGuillerme/dispRity", ref = "release")
```

```{r, eval = TRUE, echo = FALSE, message = FALSE}
## loading the package and setting up the start seed.
library(dispRity)
## Setting a random seed for repeatability
set.seed(123)
```

Note this uses the `release` branch (`r version_release`).
For the piping-hot (but potentially unstable) version, you can change the argument `ref = release` to `ref = master`.
`dispRity` depends mainly on the `ape` package and uses functions from several other packages (`ade4`, `geometry`, `grDevices`, `hypervolume`, `paleotree`, `snow`, `Claddis`, `geomorph` and `RCurl`).

## Which version do I choose? {#version}

There are always three version of the package available:
    
  * The CRAN one
  * The GitHub `release` one
  * The GitHub `master` one

The differences between the CRAN one and the GitHub `release` or `master` ones is explained just above.
For the the GitHub version, the differences are that the `release` one is more stable (i.e. more rarely modified) and the `master` one is more live one (i.e. bug fixes and new functionalities are added as they come).

If you want the latest-latest version of the package I suggest using the GitHub `master` one, especially if you recently emailed me reporting a minor bug or wanting a new functionality!
Note however that *it can happen* that the `master` version can sometimes be bugged (especially when there are major `R` and `R` packages updates), however, the status of the package state on both the `release` and the `master` version is constantly displayed on the [`README` page](https://github.com/TGuillerme/dispRity/) of the package with the nice badges displaying these different (and constantly tested) information.

## Help

If you need help with the package, hopefully the following manual will be useful.
However, parts of this package are still in development and some other parts are probably not covered.
Thus if you have suggestions or comments on on what has already been developed or will be developed, please send me an email (<guillert@tcd.ie>) or if you are a GitHub user, directly create an issue on the [GitHub page](https://github.com/TGuillerme/dispRity).

## Citations

To cite the package, this manual or some specific functionalities, you can use the following references:

The package main paper:

> Guillerme T. dispRity: A modular R package for measuring disparity. Methods Ecol Evol. 2018;9:1755–1763. [doi.org/10.1111/2041-210X.13022](https://doi.org/10.1111/2041-210X.13022).

The package manual (regularly updated!):

> Guillerme, T. & Cooper, N. (2018): dispRity manual. figshare. Preprint. [10.6084/m9.figshare.6187337.v1](https://doi.org/10.6084/m9.figshare.6187337.v1).

The time-slicing method implemented in `chrono.subsets` (unfortunately not Open Access, but you can still get a [free copy from here](https://tguillerme.github.io/papers/Guillerme&Cooper-2018-Palaeontology.pdf)):

> Guillerme, T. and Cooper, N. (2018), Time for a rethink: time sub-sampling methods in disparity-through-time analyses. Palaeontology, 61: 481-493. [doi:10.1111/pala.12364](https://onlinelibrary.wiley.com/doi/abs/10.1111/pala.12364).

Furthermore, don't forget to cite `R`:

> R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.

Bonus: you can also cite `ape` since the `dispRity` package heavily relies on it:

> Paradis E. & Schliep K. 2019. ape 5.0: an environment for modern phylogenetics and evolutionary analyses in R. Bioinformatics 35: 526-528.


### Why is it important to cite us?

Aside from how science works (if you're using a method from a specific paper, cite that specific paper to refer to that specific method), why is it important to also cite the package and the manual?

All the people involve in making the `dispRity` package happened to do it enthusiastically, freely and most amazingly without asking anything in return!
I created the package with this idea in mind and I am still sticking to it.
However, academia (the institutions and people producing science around the globe) is unfortunately not optimal at many level (some might even say "broken"): high impact papers attract big grants that attract high impact papers and big grants again, all this along with livelihood, permanent position and job security.
Unfortunately however, method development has a hard time to catch up with the current publish or perish system: constantly updating the `dispRity` package and this manual is hugely time consuming (but really fun!) and that is not even taking into account maintenance and helping users.
Although I do truly believe that this time spent doing these things modestly help the scientific endeavour, it does not contribute to our paper list!

Therefore, by citing the package and this manual, you help provide visibility to other workers and you might help them in their work!
And you directly contribute in making this project fun for all the people involved and most of all, free, updated and independent from the publish and perish system!

Thank you!
<!-- ## Associated papers -->

<!-- Here are some papers that have used the `dispRity` package: -->
